# Sample Data 

We've pulled together sample data for 715 mock students with AI/AN ancestry. This is replicated data and these are not real student `CCCID`'s. A `CCCID` code is typically three letters followed by four numbers, like `ABC1234`. We've replaced these with mock IDs that use four letters, `ABCD123`, to avoid any unintentional overlap with true students.

```{r message=FALSE, warning=FALSE}
library(tidyverse) # For data manipulation

students <- read.csv("data/studentSample.csv")
```

This sample is made up of three components: data collected from CCCApply, data for MIS/IPEDS reporting, and derived elements. 

## Application and Reporting Data

**CCCApply Data** 

```{r}
# Grab only the CCCApply columns, first 3 rows
head(
  students[c(1,4:7)], 3
)
```


**MIS/IPEDS Reporting**

Next are institutional data elements for MIS/IPEDS reporting. These should be familiar, even if the column names are a little different: 

```{r}
# Preview only the MIS/IPEDS elements, first 3 rows
head(
  students[c(1:3)], 3
)
```

## Derived Elements

Last are 16 derived elements that identify student with `any` ancestry, multiracial `mult` ancestry, or `only` a given ancestry. The latter is reserved for Hispanic/Latinx students without any ancestries outside Latin America and Spain, `onlyHispLat`. This creates a comparison group for the traditional `Hispanic/Latinx` MIS/IPEDS group.  

```{r}
# Preview only the derived elements, first 3 rows 
head(
  students[c(1, 8:23)], 3
)
```

