# UpSet Plots

UpSet plots require the package `UpSetR`. 

Please see the package documentation for introductory examples and reference manuals: 

[https://CRAN.R-project.org/package=UpSetR](https://CRAN.R-project.org/package=UpSetR)

```{r}
#install.packages("UpSetR")
library(UpSetR)
```

We'll use only the derived `any` data elements for the UpSet plots. See the section on *Sample Data* for more details. 

```{r}
# Create a reduced size data frame with just the `any` elements in it, cols 9-15. 
anyAIAN <- students[8:15] 

# Show the first three rows
head(anyAIAN, 3) 
```

UpSet plots can take a larger combination of sets and create detail rich displays. Below is a plot of all of our AI/AN students and their ancestral make-ups. 

```{r UpSet}

par(mar = c(0, 2, 2, 2))                  # Set margin size. 
upset(anyAIAN,                            # Plot our AI/AN student df
      sets = c("anyAIAN","anyAABlack",    # UpSet requires us to create 
               "anyAsian", "anyFilipino", #   a list of combinations 
               "anyPI", "anyWhite",
               "anyHispLat"), 
      nintersects = 15,                   # Show the n largest combinations
      set_size.show = T,                  # Show the counts (x-axis)
      order.by = "freq",                  # Order by most common
      sets.x.label = "Distinct Students", # Add titles 
      mainbar.y.label = "Distinct Students", 
      text.scale = 
        c(1.3, 1.3, 1, 1, 1.3, 1.5)       # Adjust text sizes for plot
)
```
The main bar plot details the number of student with a given combination or `intersection`. The dots below the main bar plot are the ancestry combinations. In this figure, the most common ancestry combination is White and AI/AN. 

Notice that every student represented in the plot has AI/AN ancestry. The smaller, horizontal bar plot to the left of each ancestry counts the number of students who share that ancestry. UpSet refers to these groupings as `sets` and the counts as `set frequencies`. 

With 715 total AI/AN students, 453 have some White ancestry, and 291 with some Hispanic/Latinx ancestry. After White and Hispanic ancestry there is a large drop to students that share some African American/Black, Asian, Filipino or Pacific Islander ancestry. But these small groups aren't lost like they were in the Euler plots!


The figure above is limited to the top 15 combination. There aren't any tie-breaker rules if, like in our case, the 15th and 16th combination are the same size. To expand the figure out for all values, omit the `nintersects` variable in the `upset()` function.

```{r UpSet2}
upset(anyAIAN,                            # Plot our AI/AN student df
      sets = c("anyAIAN","anyAABlack",    # UpSet requires us to create 
               "anyAsian", "anyFilipino", #   a list of combinations 
               "anyPI", "anyWhite",
               "anyHispLat"), 
      #nintersects = 15,                  # Omit n largest combo filter
      set_size.show = T,                  # Show the counts (x-axis)
      order.by = "freq",                  # Order by most common
      sets.x.label = "Distinct Students", # Add titles 
      mainbar.y.label = "Distinct Students", 
      text.scale = 
        c(1.3, 1.3, 1, 1, 1.3, 1.5)       # Adjust text sizes for plot
)
```

Queries are an interesting addition to UpSet plots. These can get complicated in a hurry, though. The vignette can be found here: [Querying the Data](https://cran.r-project.org/web/packages/UpSetR/vignettes/queries.html)


```{r}
upset(anyAIAN,                            # Plot our AI/AN student df
      sets = c("anyAIAN","anyAABlack",    # UpSet requires us to create 
               "anyAsian", "anyFilipino", #   a list of combinations 
               "anyPI", "anyWhite",
               "anyHispLat"), 
      set_size.show = T,                   # Show the counts (x-axis)
      order.by = "freq",                   # Order by most common
      cutoff = 3, 
      sets.x.label = "Distinct Students",  # Add titles 
      mainbar.y.label = "Distinct Students", 
      text.scale = 
        c(1.3, 1.3, 1, 1, 1.3, 1.5),       # Adjust text sizes for plot
      
      # Query! Must be a list of lists. 
      queries = list(
                  list(query = intersects, # Highlight intersections
                  params = "anyAIAN",      # Showing our MIS AI/AN
                  color = "goldenrod",   
                  active = T)
      )
)
```